---
title: "Discussion5_Summarizing Dynamic Model Output"
format: revealjs
theme: solarized
resources: ["img/"]
css: ["slides.css"]
editor: visual
---

## Adding Carrying Capacity

## R code with carrying capacity {.scrollable}

```{r exampleode}
library(deSolve)
library(tidyverse)
library(here)
library(ggplot2)

source(here("R/dexppop_play.R"))

dexppop_play

# Create parameter list
initalrabbits <- 2
years <- seq(from = 1, to = 100, by = 2)
newparms <- list(r = 0.03, carry_capacity = 300)

# Apply solver
results <- ode(initalrabbits, years, dexppop_play, newparms)
head(results)

# Add more meaningful names
colnames(results) <- c("year", "P")
```

## Plot

```{r, plotodep}
# Plot
ggplot(as.data.frame(results), aes(year, P)) +
  geom_point() +
  labs(x = "years", y = "Population")
```

## Try again with different parameters {.scrollable}

```{r exampleode2}
# Same initial condtions
initalrabbits <- 2
years <- seq(from = 1, to = 100, by = 2)

# Try again with different parameters
alternativeparms <- list(r = 0.04, carry_capacity = 500)
results2 <- ode(initalrabbits, years, dexppop_play, alternativeparms)

# look at results
head(results2)
colnames(results2) <- c("year", "P_parm2")

# Plot
ggplot(as.data.frame(results2), aes(year, P_parm2)) +
  geom_point() +
  labs(y = "Population", x="years")

# Compare by combining into a single data frame
both <- inner_join(as.data.frame(results), as.data.frame(results2))

# both_p <- both %>% gather(key = "model", value = "P", -year)
both_p <- both %>% pivot_longer(-year, names_to = "model", values_to = "P")
```

## Plot

```{r, plotb}
# Plot
ggplot(both_p, aes(year, P, col = model)) +
  geom_point() +
  labs(x = "years", y = "Population")
```

## Extracting information from space-time results {.scrollable}

In class we've talking about summarizing results from dynamic models

-   pictures can be hard to interpret

-   summarizing over one of the dimensions (either space or time) can help

-   looking at a single trajectory through time

-   looking at spatial variation for one point in time

-   looking at spatial variation for multiple points in time

## Extracting from our diffusion model {.scrollable}

```{r}
source(here("R/diffusion.R"))
resfast_fixtime <- diff1(initialC = 100, nx = 10, dx = 0.5, 
                         nt = 100, dt = 0.5, D = 0.2, area = 1)


filled.contour(resfast_fixtime$conc, xlab = "time", 
               ylab = "Distance Along Path", 
               main = "Concentration through time and space")
```

# Summarizing Model Output

-   Summarize relative to what you care about

Lets say this was a pollutant injected into a stream...

-   lets say you have a vulnerable habit located 3 spatial units from the injection point

-   lets see what its pollutant exposure looks like.\

```{r one option}
# View(resfast_fixtime$conc)

# graph a single point in space through time
# single column (time)
plot(resfast_fixtime$conc[ , 3], 
     ylab = "Concentration at 2 spatial time units from origin (1+2)",
     xlab = "Time step")


```

-   what if we were concerned about all locations

```{r all locations}
# plot all trajectories
# add a time column to concentration data frame and transform for plotting
resl <- as.data.frame(resfast_fixtime$conc) %>%
  mutate(time = seq(from = 1, to = 100)) %>%
  pivot_longer(-time, names_to = "distance", values_to = "conc")

ggplot(resl, aes(time, conc, col = distance)) +
  geom_line()

# plot all places at each point in time
ggplot(resl, aes(time, conc, group = time)) +
  geom_boxplot()
```

# Spatial summaries

-   interested in how quickly the pollutant evens out through space

```{r spatial summaries}
# use apply to calculate the spatial variation for each row (e.g for each time point)
cvar <- resfast_fixtime$conc %>% 
  apply(1, var)
cmean <- resfast_fixtime$conc %>% 
  apply(1, mean)

spatial_aver <- cbind.data.frame(cvar, cmean, time = seq(from = 1, to = 100))
length(cvar)

# notice its the same as the number of time units (nt) used above
# plot spatial variation through time
ggplot(spatial_aver, aes(time, cvar)) +
  geom_line() +
  labs(y = "Spatial Variation")

# plot coefficient of variation (so standard deviation divided by the mean)
ggplot(spatial_aver, aes(time, 100 * sqrt(cvar) / cmean)) +
  geom_line() +
  labs(y = "COV (as percent")
```

## it is helpful to turn the summary into a function that you can apply to the dynamic models output

## In class exercise {.scrollable}

For population model - write a function that summarizes something about the population over time

```{r}
# df A data frame containing the population records
# year_col Character string of the column name for years
# pop_col Character string of the column name for population counts
# Return a summary data frame with total growth, avg change, and peak years
summarize_population <- function(df, year_col, pop_col) {
  
  # Order the data by year
  df <- df[order(df[[year_col]]), ]
  
  # Calculate simple differences
  pops <- df[[pop_col]]
  growth <- diff(pops)
  
  # Return a list or data frame of "Big Picture" stats
  data.frame(Initial_Pop = pops[1],
             Final_Pop   = pops[length(pops)],
             Total_Change = pops[length(pops)] - pops[1],
             Peak_Growth_Value = max(growth, na.rm = TRUE))

}

# Example Usage:
# pop_df <- data.frame(Year = 2020:2024, Pop = c(100, 105, 115, 112, 120))
# summarize_population(pop_df, "Year", "Pop")
```
